Global patterns of forest autotrophic carbon fluxes

### 2020-06-05 manuscript/manuscript draft.Rmd
standardize variable symbols in main text
#55

All major carbon fluxes decreased linearly with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes--particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). The FACF that were most poorly predicted by latitude were $BNPP_{root-fine}$ ($R^2$=0.17) and $ANPP_{woody.stem}$ ($R^2$=0.18). For all FACF, the relationship with latitude was best predicted by the first-order linear model.

@@

In general, smaller component fluxes summed approximately to larger fluxes across the latitudinal gradient (Fig. 2). That is, modelled estimates of $GPP$, generated from the sum of $NPP$ and $R_{auto}$; $NPP$, generated from the sum of $ANPP$ and $BNPP$; and $ANPP$, generated from the sum of $ANPP_{foliage}$ and $ANPP_{woody-stem}$, fell completely within the confidence intervals of the regressions of field estimates of $GPP$, $NPP$, and $ANPP$, respectively.

@@

All FACF increased linearly with MAT, and we found no support for a saturation point of FACF with MAT (all p<0.05; Figs. 3-4, S4-S5, Table S2). As with latitude, MAT tended to explain more variation in the larger FACF ($GPP$, $NPP$, $ANPP$, $R_{auto}$) and $ANPP_{foliage}$ (all $R^2$> 0.4) than in subsidiary and belowground fluxes ($ANPP_{woody-stem}$, $R_{root}$, $BNPP_{root-fine}$; all $R^2$< 0.25).

MAP was a significant (p<0.05) predictor of all FACF except $ANPP_{woody-stem}$ (Figs. 4a, S4-S5; Table S2). However, it explained little variation: with the exception of $R_{auto}$, MAP explained at most 37% of variation in FACF. For the majority of FACF, a polynomial model **BBL: of what order?** provided the best fit. FACF generally increased with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which they started to decrease (Figs. 4, S4-S5). The notable exception to this was GPP: the model indicated that GPP continued to increase with precipitation up to measures of at least 5000mm annually (p<0.0001, $R^2$ = 0.33. Data above this point were not available, but the model trend suggested a saturation point around 5000mm MAP.

There was a significant additive effect of MAT and MAP on $GPP$, $ANPP$ and $R_{auto}$ (Fig. 3, Table S3). Accounting for MAT, MAP had a substantial positive effect on $GPP$ and $R_{auto}$ and a small negative effect on $ANPP$. There was a significant interactive effect between MAT and MAP for $NPP$ and $ANPP_{woody-stem}$ (Fig. 3, Table S3). The interaction was negative for $NPP$ and positive for $ANPP_{woody-stem}$. For $ANPP_{foliage}$, $BNPP$, $BNPP_{root-fine}$, and $R_{auto-root}$, MAP did not have a signficant effect when accounting for MAT (Fig. 3, Table S3). For the variables which showed a significant interactive or additive effect between MAT and MAP, no other climate variable, in combination with MAT, significantly improved on that model. {**need to confirm this given changes in MAT MAP results (or you could just drop the sentence.)**}
**BBL: this paragraph is pretty dense. Perhaps distill to a summary sentence and move rest to SI?**

@@

We found a significant relationship between C flux and potential evapotranspiration for all FACF. $ANPP_{foliage}$, $BNPP_{root-fine}$ and $R_{root}$ increased linearly with PET; however, all other fluxes showed a polynomial relationship with PET (Fig. 4c, S4-5; Table S2). We found strong evidence for a saturation point or peak with PET: FACF tended to increase at values below 1000mm, before saturating between 1200 and 1700mm. There was also evidence that FACF begin to decrease at values above 1800mm PET.

Vapour pressure deficit was a significant predictor of C flux for all FACF. $BNPP_{root-fine}$ showed a linear relationship with vapour pressure deficit ($R^2$ = 0.07, p<0.05), but all other fluxes showed a polynomial relationship (Figs. 4d, S4-5; Table S2). FACF initially increased with vapour pressure deficit, before saturating at around 0.8 kPa, after wich point they began to decrease.

@@

Temperature seasonality was a significant predictor of FACF. $GPP$, $NPP$, $ANPP$, and $R_{root}$ exhibited a polynomial relationship with seasonality (all p<0.05; Figs. 4e, S6-7; Table S2). $ANPP_{foliage}$, $ANPP_{woody-stem}$ and $R_{auto}$ decreased linearly with temperature seasonality (all p<0.05; Figs. 4e, S6-S7; Table S2). Temperature seasonality was strongly correlated with annual temperature range, which was likewise a similarly strong predictor of FACF (Table S2). FACF were highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower. **BBL: perhaps put this into an ecosystem context; what are these? Aseasonal subtropical places?**

@@

When FACF were standardized by growing season length, correlations with growing season climate--including temperature, precipitation, solar radiation, and PET--were generally weak (Figs. S8-S9). Speficifally, $ANPP$ increased with growing season temperature ($R^2$ = 0.10, p<0.001) and precipitation ($R^2$ = 0.04, p<0.05). Similarly, $ANPP_{foliage}$ increased slightly with growing season temperature ($R^2$ = 0.16, p<0.01) and precipitation ($R^2$ = 0.09, p<0.05). Growing season solar radiation had a positive influence on $GPP$ ($R^2$ = 0.21, p<0.001), $NPP$ ($R^2$ = 0.21, p<0.001), $BNPP$ ($R^2$ = 0.16, p<0.001) and $BNPP_{fine.root}$ ($R^2$ = 0.12, p<0.01). Growing season PET had a positive influence on $GPP$ ($R^2$ = 0.15, p<0.01), $NPP$ ($R^2$ = 0.18, p<0.01), $BNPP$ ($R^2$ = 0.23, p<0.0001), $BNPP_{fine.root}$ ($R^2$ = 0.11, p<0.05), and $ANPP_{woody-stem}$ ($R^2$ = 0.06, p<0.05). {**Becky, please verify/ edit the following:** There were no other signifcant correlations between growing season length-standardized FACF (9 variables in Table 2) and growing season climate (**which variables?**)}.

@@

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine FACF analyzed, C cycling decreases continually with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. FACF generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (*e.g.*, $ANPP_{woody}$, $BNPP_{fine.root}$, $R_{auto-root}$; Fig. 2; Table S2).

@@

We show that FACF are broadly consistent in their responses to climate drivers on the global scale, with at most modest trends in C allocation among the variable pairs tested (Figs. 2, S3). 
This parallels the observation that C allocation across multiple FACF varies little with respect to climate along a steep tropical elevational gradient [@malhi_variation_2017; but see @moser_elevation_2011], 
and is not surprising given that trees face universal constraints in carbon allocation (**REFS**). 
The one trend in allocation that emerged from our analysis was a tendency for greater C allocation aboveground--and to foliage in particular--in warm tropical climates (Fig. S3). This is consistent with observations of increasing allocation to roots with declining temperature across a tropical elevational gradient [@moser_elevation_201], 
and with observations and theory predicting relatively higher belowground carbon allocation at higher latitudes [@gill_belowground_2016].
It is also consistent with findings that as forest productivity increases, proportionaly more carbon is allocated to $ANPP_{foliage}$ relative to $BNPP_{root-fine}$ (*Chen et al. 2019; doi: 10.1111/ele.13193*) or $ANPP_{woody}$ [@hofhansl_new_2015]. **( check / comment on  @litton_carbon_2007 )**

@@

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$) than in subsidiary fluxes ($BNPP_{fine.root}$, $R_{auto-root}$, $ANPP_{woody-stem}$) (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (*Clark et al. 2001 Ecological Applications*). However, methodological variation and uncertainty affect all of fluxes considered here, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

>>>>

All major carbon fluxes decreased linearly with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes--particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). The FACF that were most poorly predicted by latitude were $BNPP_{fine.root}$ ($R^2$=0.17) and $ANPP_{woody.stem}$ ($R^2$=0.18). For all FACF, the relationship with latitude was best predicted by the first-order linear model.

@@

In general, smaller component fluxes summed approximately to larger fluxes across the latitudinal gradient (Fig. 2). That is, modelled estimates of $GPP$, generated from the sum of $NPP$ and $R_{auto}$; $NPP$, generated from the sum of $ANPP$ and $BNPP$; and $ANPP$, generated from the sum of $ANPP_{foliage}$ and $ANPP_{stem}$, fell completely within the confidence intervals of the regressions of field estimates of $GPP$, $NPP$, and $ANPP$, respectively.

@@

All FACF increased linearly with MAT, and we found no support for a saturation point of FACF with MAT (all p<0.05; Figs. 3-4, S4-S5, Table S2). As with latitude, MAT tended to explain more variation in the larger FACF ($GPP$, $NPP$, $ANPP$, $R_{auto}$) and $ANPP_{foliage}$ (all $R^2$> 0.4) than in subsidiary and belowground fluxes ($ANPP_{stem}$, $R_{root}$, $BNPP_{fine.root}$; all $R^2$< 0.25).

MAP was a significant (p<0.05) predictor of all FACF except $ANPP_{stem}$ (Figs. 4a, S4-S5; Table S2). However, it explained little variation: with the exception of $R_{auto}$, MAP explained at most 37% of variation in FACF. For the majority of FACF, a polynomial model **BBL: of what order?** provided the best fit. FACF generally increased with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which they started to decrease (Figs. 4, S4-S5). The notable exception to this was GPP: the model indicated that GPP continued to increase with precipitation up to measures of at least 5000mm annually (p<0.0001, $R^2$ = 0.33. Data above this point were not available, but the model trend suggested a saturation point around 5000mm MAP.

There was a significant additive effect of MAT and MAP on $GPP$, $ANPP$ and $R_{auto}$ (Fig. 3, Table S3). Accounting for MAT, MAP had a substantial positive effect on $GPP$ and $R_{auto}$ and a small negative effect on $ANPP$. There was a significant interactive effect between MAT and MAP for $NPP$ and $ANPP_{stem}$ (Fig. 3, Table S3). The interaction was negative for $NPP$ and positive for $ANPP_{stem}$. For $ANPP_{foliage}$, $BNPP$, $BNPP_{fine.root}$, and $R_{root}$, MAP did not have a signficant effect when accounting for MAT (Fig. 3, Table S3). For the variables which showed a significant interactive or additive effect between MAT and MAP, no other climate variable, in combination with MAT, significantly improved on that model. {**need to confirm this given changes in MAT MAP results (or you could just drop the sentence.)**}
**BBL: this paragraph is pretty dense. Perhaps distill to a summary sentence and move rest to SI?**

@@

We found a significant relationship between C flux and potential evapotranspiration for all FACF. $ANPP_{foliage}$, $BNPP_{fine.root}$ and $R_{root}$ increased linearly with PET; however, all other fluxes showed a polynomial relationship with PET (Fig. 4c, S4-5; Table S2). We found strong evidence for a saturation point or peak with PET: FACF tended to increase at values below 1000mm, before saturating between 1200 and 1700mm. There was also evidence that FACF begin to decrease at values above 1800mm PET.

Vapour pressure deficit was a significant predictor of C flux for all FACF. $BNPP_{fine.root}$ showed a linear relationship with vapour pressure deficit ($R^2$ = 0.07, p<0.05), but all other fluxes showed a polynomial relationship (Figs. 4d, S4-5; Table S2). FACF initially increased with vapour pressure deficit, before saturating at around 0.8 kPa, after wich point they began to decrease.

@@

Temperature seasonality was a significant predictor of FACF. $GPP$, $NPP$, $ANPP$, and $R_{root}$ exhibited a polynomial relationship with seasonality (all p<0.05; Figs. 4e, S6-7; Table S2). $ANPP_{foliage}$, $ANPP_{stem}$ and $R_{auto}$ decreased linearly with temperature seasonality (all p<0.05; Figs. 4e, S6-S7; Table S2). Temperature seasonality was strongly correlated with annual temperature range, which was likewise a similarly strong predictor of FACF (Table S2). FACF were highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower. **BBL: perhaps put this into an ecosystem context; what are these? Aseasonal subtropical places?**

@@

When FACF were standardized by growing season length, correlations with growing season climate--including temperature, precipitation, solar radiation, and PET--were generally weak (Figs. S8-S9). Speficifally, $ANPP$ increased with growing season temperature ($R^2$ = 0.10, p<0.001) and precipitation ($R^2$ = 0.04, p<0.05). Similarly, $ANPP_{foliage}$ increased slightly with growing season temperature ($R^2$ = 0.16, p<0.01) and precipitation ($R^2$ = 0.09, p<0.05). Growing season solar radiation had a positive influence on $GPP$ ($R^2$ = 0.21, p<0.001), $NPP$ ($R^2$ = 0.21, p<0.001), $BNPP$ ($R^2$ = 0.16, p<0.001) and $BNPP_{fine.root}$ ($R^2$ = 0.12, p<0.01). Growing season PET had a positive influence on $GPP$ ($R^2$ = 0.15, p<0.01), $NPP$ ($R^2$ = 0.18, p<0.01), $BNPP$ ($R^2$ = 0.23, p<0.0001), $BNPP_{fine.root}$ ($R^2$ = 0.11, p<0.05), and $ANPP_{stem}$ ($R^2$ = 0.06, p<0.05). {**Becky, please verify/ edit the following:** There were no other signifcant correlations between growing season length-standardized FACF (9 variables in Table 2) and growing season climate (**which variables?**)}.

@@

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine FACF analyzed, C cycling decreases continually with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. FACF generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (*e.g.*, $ANPP_{woody}$, $BNPP_{fine.root}$, $R_{root}$; Fig. 2; Table S2).

@@

We show that FACF are broadly consistent in their responses to climate drivers on the global scale, with at most modest trends in C allocation among the variable pairs tested (Figs. 2, S3). 
This parallels the observation that C allocation across multiple FACF varies little with respect to climate along a steep tropical elevational gradient [@malhi_variation_2017; but see @moser_elevation_2011], 
and is not surprising given that trees face universal constraints in carbon allocation (**REFS**). 
The one trend in allocation that emerged from our analysis was a tendency for greater C allocation aboveground--and to foliage in particular--in warm tropical climates (Fig. S3). This is consistent with observations of increasing allocation to roots with declining temperature across a tropical elevational gradient [@moser_elevation_201], 
and with observations and theory predicting relatively higher belowground carbon allocation at higher latitudes [@gill_belowground_2016].
It is also consistent with findings that as forest productivity increases, proportionaly more carbon is allocated to $ANPP_{foliage}$ relative to $BNPP_{fine.root}$ (*Chen et al. 2019; doi: 10.1111/ele.13193*) or $ANPP_{woody}$ [@hofhansl_new_2015]. **( check / comment on  @litton_carbon_2007 )**

@@

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$) than in subsidiary fluxes ($BNPP_{fine.root}$, $R_{root}$, $ANPP_{stem}$) (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{fine.root}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (*Clark et al. 2001 Ecological Applications*). However, methodological variation and uncertainty affect all of fluxes considered here, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).
